Reversible Regime Change: Climate-driven phytoplankton community shifts in the Cariaco Basin, Venezuela

### 2025-07-28 agujournaltemplate.tex
Final edits, acknowledgements etc.

\subsection{Meteorological and climate data}
    %from Taylor et al. 2012
    In addition to oceanographic environmental data, we also examine meteorological and climatic data that are known to influence ocean conditions and marine ecosystems. Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}. We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as "wind speed" throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate of total precipitation, comprising all water that falls on Earth's surface as rain or snow.

@@

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using tidyverse packages \cite{wickham_welcome_2019}. All the codes we produced for data processing and data analysis are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 \meter for further analysis. Furthermore, the sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}. For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated and integrated across depth, and finally grouped by genus via the sum of interpolated counts.

\subsection{Data analysis}    
    In the analysis, we set out to find if there is a structure in the distribution and temporal evolution of diversity metrics and the community composition, as well as to identify potential bottom-up drivers of this change. 
    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where $\mu$ is the mean across all available data points). Z scores from 1995, 2016, and 2017 were not shown because insufficient coverage (i.e. more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.

    Cell counts of phytoplankton per functional group were assembled per monthly sampling, summed over all individual interpolated counts at the genus or species level, and finally averaged over a year. Nanoflagellate counts were reported as such in the data and not further identified. To characterize phytoplankton community structure, we calculated genus richness, Shannon diversity (H'), and Pielou’s evenness (J') to assess taxonomic variety, community complexity, and the distribution of abundances among genera, respectively. We calculated genus richness as the sum of all identified genera present in a monthly sample. Genus richness provides a lower taxonomic resolution than species richness, but creates a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}. The Shannon index was calculated as $H' = -\sum_i p_i \log_{e} p_i$ and Pielou's evenness as $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling. Diversity metrics were calculated for each monthly sample and averaged throughout the year.

@@

%!
\acknowledgments
Thank you to all who have helped, communicated, provided data. Claudia Benitez-Nelson, Digna Rueda-Roa, James Pinckney.

>>>>

\subsection{Meteorological and climate data}
    %from Taylor et al. 2012
    In addition to the in situ data, we also examine meteorological and climatic data that are known to influence ocean conditions and marine ecosystems. Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}. We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as "wind speed" throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate of total precipitation, comprising all water that falls on Earth's surface as rain or snow.

@@

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using tidyverse packages \cite{wickham_welcome_2019}. All the codes we produced for data processing and data analysis are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to \qty{100}{\meter} and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis. Furthermore, the sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}. For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated and integrated across depth, and finally grouped by genus via the sum of interpolated counts.

\subsection{Data analysis}    
    In the analysis, we set out to find if there is a structure in the distribution and temporal evolution of diversity metrics and the community composition, as well as to identify potential bottom-up drivers of this change. 
    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where $x$ is the yearly mean, $\mu$ is the mean and $\sigma$ is the standard deviation of all yearly means). Data from 1995, 2016, and 2017 were not included because insufficient coverage (i.e. more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.

    Cell counts of phytoplankton per functional group were assembled per monthly sampling, summed over all individual interpolated counts at the genus or species level, and finally averaged over a year. Nanoflagellate counts were reported as such in the data and not further identified. To characterize phytoplankton community structure, we calculated genus richness, Shannon diversity (H'), and Pielou’s evenness (J') to assess taxonomic variety, community complexity, and the distribution of abundances among genera, respectively. We calculated genus richness as the sum of all identified genera present in a monthly sample. Genus richness provides a lower taxonomic resolution than species richness, but creates a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}. The Shannon index was calculated as $H' = -\sum_i [p_i \log_{e}(p_i)]$ and Pielou's evenness as $J' = -\sum_i[ p_i \log_{e}( p_i )/\log_{e}(H')]$, where $p_i$ represents the number of genera observed for each interpolated and integrated monthly sampling. Diversity metrics were calculated for each monthly sample and averaged throughout the year.

@@

%!
\acknowledgments
We thank Digna Rueda-Roa, Claudia Benitez-Nelson, and James Pinckney for their valuable input during the early stages of this research.
